---
title: "r-week-8-hw-eeb177.Rmd"
output: md_document
---
# Canids

```{r}
#this will get the data from the csv file we made from python
library(ggplot2)
setwd("/home/eeb177-student/Desktop/eeb-177/homework/canids")

canids <- read.csv("/home/eeb177-student/Desktop/eeb-177/homework/canids/canid-ranges.csv", header =F, as.is = T)

names(canids) <- c("genus", "species", "minage", "maxage")
head(canids)
```

```{r}
canid_occ <- ggplot(canids, aes( species, ymin = maxage, ymax=minage, colour = genus))
canid_occ <- canid_occ + geom_linerange()
canid_occ <- canid_occ + theme(legend.position="none") # this will get rid of the legend on the graph
canid_occ <- canid_occ + coord_flip() # this will flip the axis to maket he graph go the other direction
canid_occ <- canid_occ +  theme(axis.text.y = element_text(size=3)) # makes the text a more readable size
canid_occ <- canid_occ + theme(axis.ticks.y=element_blank()) # removes the tick marks that are present on the y axis
canid_occ <- canid_occ + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) # makes the graph fit more by removing the excess white space
canid_occ <- canid_occ + labs(title = "Canid Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))  # adding in the title and axis labels to the graph
canid_occ
```

```{r}
ggsave(filename = "canid-occ.pdf", plot = canid_occ)
```

1. What taxonomic group is being depicted in the graph?
```{r}
print("Canids")
```
 
2. How many unique fossil species are in this family?
```{r}
cat("Unique species:", nrow(canids))
```

3. How many unique fossil genera are in this family?
```{r}
genera <- unique(canids$genus)
cat("Number of Genera:", length(genera))
```

4. What species has the longest genera?
```{r}
# easier to make a new column that will calculate the ranges and then call on the row that has the maximum range
canids$length <- canids$maxage - canids$minage
max_length_species <- which.max(canids$length) 
cat(canids[max_length_species,2], canids[max_length_species,5], "Million years")
```

# Heteromyidae

```{r}
#this will get the data from the csv file we made from python
library(ggplot2)
setwd("/home/eeb177-student/Desktop/eeb-177/homework/canids")

rats <- read.csv("/home/eeb177-student/Desktop/eeb-177/homework/canids/rat-ranges.csv", header =F, as.is = T)

names(rats) <- c("genus", "species", "minage", "maxage")
head(rats)
```

```{r}
rats_occ <- ggplot(rats, aes( species, ymin = maxage, ymax=minage, colour = genus))
rats_occ <- rats_occ + geom_linerange()
rats_occ <- rats_occ + theme(legend.position="none") # this will get rid of the legend on the graph
rats_occ <- rats_occ + coord_flip() # this will flip the axis to maket he graph go the other direction
rats_occ <- rats_occ +  theme(axis.text.y = element_text(size=3)) # makes the text a more readable size
rats_occ <- rats_occ + theme(axis.ticks.y=element_blank()) # removes the tick marks that are present on the y axis
rats_occ <- rats_occ + scale_y_continuous(limits=c(0, 35), expand = c(0, 0), breaks=c(0, 5, 10, 15, 20, 25, 30, 35, 40)) # makes the graph fit more by removing the excess white space
rats_occ <- rats_occ + labs(title = "Heteromyidae Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))  # adding in the title and axis labels to the graph
rats_occ
```

```{r}
ggsave(filename = "rat-occ.pdf", plot = rats_occ)
```

1. What taxonomic group is being depicted in the graph?
```{r}
print("Heteromyidae")
```

2. How many unique fossil species are in this family?
```{r}
cat("Unique species:", nrow(rats))
```

3. How many unique fossil genera are in this family?
```{r}
genera_rat <- unique(rats$genus)
cat("Number of Genera:", length(genera_rat))
```

4. What species has the longest genera?
```{r}
# easier to make a new column that will calculate the ranges and then call on the row that has the maximum range
rats$length <- rats$maxage - rats$minage
max_length_species <- which.max(rats$length) 
cat(rats[max_length_species,2], rats[max_length_species,5], "Million years")
```

# Part 2: Data Frames

```{r}
cats <- read.csv(file = "/home/eeb177-student/Desktop/eeb-177/Class-Work/data/feline-data.csv")
age <- c(4,5,8)
cats <- cbind(cats, age)
```

```{r}
newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats, newRow)
```

```{r}
levels(cats$coat)
levels(cats$coat) <- c(levels(cats$coat), 'tortoiseshell')
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9))
cats
```

```{r}
str(cats)
```

```{r}
cats$coat <- as.character(cats$coat)
str(cats)
```

```{r}
cats[-4,]
```

```{r}
cats <- na.omit(cats)
```

```{r}
cats <- rbind(cats, cats)
cats
rownames(cats) <- NULL
cats
```

## Challenge 1

```{r}
df <- data.frame(first = c('Grace'),
                 last = c('Hopper'),
                 lucky_number = c(0),
                 stringsAsFactors = FALSE)
df <- rbind(df, list('Marie', 'Curie', 238) )
df <- cbind(df, coffeetime = c(TRUE,TRUE))
df
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")

```

```{r}
str(gapminder)
```

```{r}
typeof(gapminder$year)
```

```{r}
typeof(gapminder$country)
```

```{r}
str(gapminder$country)
```

```{r}
length(gapminder)
```

```{r}
typeof(gapminder)
```

```{r}
dim(gapminder)
```

```{r}
colnames(gapminder)
```
 
```{r}
head(gapminder)
```
 
## Challenge 2 
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")
#source(file = "~/scripts/load-gapminder.R")
```
 
## Challenge 3

gapminder is a data frame with columns that read: 1704 rows and 6 columns. Country and Continent are factors. Year is an integer vector. pop, lifeExp and gdpPercap are numeric vectors.

# Part 2: Subsetting Data

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```
 
```{r}
x[1]
x[4]
x[c(1,3)]
x[1:4]
```
 
```{r}
x[c(1,1,3)]
x[6]
x[0]
```
 
```{r}
x[-2]
x[c(-1, -5)]
```
 
```{r}
x <- x[-4]
x
```
 
## Challenge 1

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)

x[2:4]
x[-c(1,5)]
x[c("b", "c", "d")]
x[c(2,3,4)]
```

```{r}
x[c("a", "c")]
```

```{r}
x[-which(names(x) == "a")]
names(x) == "a"
which(names(x) == "a")
x[-which(names(x) %in% c("a", "c"))]
```

## Challenge 2

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```
 
```{r}
x[-which(names(x) == "g")]
```
 
if there are no TRUE values passed to which, an empty vector is returned
integer() is an empty vector

```{r}
names(x) <- c('a', 'a', 'a')
x['a']
x[which(names(x) == 'a')]
```
 
```{r}
names(x) == c('a', 'c', 'e')
```

```{r} 
x[c(TRUE, TRUE, FALSE, FALSE)]
x[c(TRUE, FALSE)]
x[x > 7]
```

## Challenge 3

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)

x_subset <- x[x<7 & x>4]
print(x_subset)
```

```{r}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
```

```{r}
f[f %in% c("b", "c")]
f[1:3]
f[-3]
```

```{r}
set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]
```

```{r}
m[, c(3,4)]
```

```{r}
m[3,]
m[3, , drop=FALSE]
```

```{r}
m[5]
```
 
```{r}
matrix(1:6, nrow=2, ncol=3)
```

```{r}
matrix(1:6, nrow=2, ncol=3, byrow=TRUE)
```
 
## Challenge 4

```{r}
m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

 m[2,c(4,5)] will extract the values 11 and 14.
 
```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist[1]
```

```{r}
xlist[1:2]
```

```{r}
xlist[[1]]
```

```{r}
xlist[["a"]]
```

```{r}
xlist$data
```

## Challenge 5

```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist[[2]][2]
```

## Challenge 6

```{r}
mod <- aov(pop ~ lifeExp, data=gapminder)
attributes(mod)
mod$df.residual
```

```{r}
head(gapminder[3])
```

```{r}
head(gapminder[["lifeExp"]])
```
 
```{r}
head(gapminder$year)
```
 
```{r}
gapminder[1:3,]
```
 
```{r}
gapminder[3,]
```
 
## Challenge 7

```{r}
gapminder[gapminder$year == 1957,]
```
 
```{r}
gapminder[,-c(1:4)]
```

```{r}
gapminder[gapminder$lifeExp > 80,]
```

```{r}
gapminder[1, c(4, 5)]
```

```{r}
gapminder[gapminder$year == 2002 | gapminder$year == 2007,]
gapminder[gapminder$year %in% c(2002, 2007),]
```

## Challenge 8 

gapminder is a data.frame so needs to be subsetted on two dimensions. gapminder[1:20, ] subsets the data to give the first 20 rows and all columns.
```{r}
gapminder_small <- gapminder[c(1:9, 19:23),]
```

# Part 2: Creating Publication Quality Graphics

```{r}
library("ggplot2")
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

## Challenge 1

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp)) + geom_point()
```

## Challenge 2

```{r}
ggplot(data = gapminder, aes(x = year, y = lifeExp, color=continent)) +
  geom_point()
```

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```

## Challenge 3

The lines get drawn over the points.
```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
 geom_point() + geom_line(aes(color=continent))
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color=continent)) +
  geom_point()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5) + scale_x_log10()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm")
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

## Challenge 4a

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
 geom_point(size=3, color="orange") + scale_x_log10() +
 geom_smooth(method="lm", size=1.5)
```

## Challenge 4b

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
geom_point(size=3, pch=17) + scale_x_log10() +
geom_smooth(method="lm", size=1.5)
```

```{r}
starts.with <- substr(gapminder$country, start = 1, stop = 1)
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```

```{r}
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 1") +
  scale_colour_discrete(name="Continent") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

## Challenge 5

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, fill=continent)) +
 geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
```


















 